---
title: "Analysis of Damage Caused by Weather Phenomena Using Open Data"
author: Ivan Lysiuchenko
output: html_document
---

## Synopsis

Weather phenomena and their consequences play an extremely important role in human life.
Many of them are able to disrupt normal life, cause death or injuries, damage property and critical infrastructure.
This research is based on open data from the US National Oceanic and Atmospheric Administration (NOAA).
The aim was to detect the weather phenomena which are most harmful for people's life and health, as well as
those causing most damage to economy. To achieve the mentioned goal, the total impact on the US population and economy 
was computed over the monitoring period starting in January, 1950 and ending in November, 2011.

## Data Processing

We start by reading the data from the provided compressed CSV file.

```{r}
dataFileName <- "repdata_data_StormData.csv.bz2"
storms <- read.csv(dataFileName)
```
Before we are able to do research to answer our questions, some preprocessing
is needed.
The values of damage (PROPDMG and CROPDMG) are given in a weird form, represented as a value and exponent.
The exponent is either a number or as a character ('0' means 10<sup>0</sup>=1, '4' means 10<sup>4</sup>, 'K' means 10<sup>3</sup> and so on).
Sometimes the exponent fields are used to mark missing estimates (like '-' or '?').
To make it easy to work with damage values we compute the total damage ('prop' plus 'crop')
adding an extra numeric column called DAMAGENUMERIC, as well as two columns for the two individual components of economic damage,
PROPDMGNUMERIC and CROPDMGNUMERIC.

```{r}
exponentToMultiplier <- function(exponent)
{
    expCode <- toupper(exponent)
    res <- expCode
    res[is.na(expCode)] <- 0
    res[expCode %in% c("-", "?", "NA")] <- 0
    res[expCode %in% 0:8] <- 10 ^ as.integer(expCode[expCode %in% 0:8])
    res[expCode %in% c(" ", "", "+")] <- 1
    res[expCode == "H"] <- 100
    res[expCode == "K"] <- 1000
    res[expCode == "M"] <- 1e6
    res[expCode == "B"] <- 1e9
    
    as.numeric(res)
}

library(dplyr)
storms <- mutate(storms, DAMAGENUMERIC = PROPDMG * exponentToMultiplier(PROPDMGEXP) +
                     CROPDMG * exponentToMultiplier(CROPDMGEXP), 
                 PROPDMGNUMERIC = PROPDMG * exponentToMultiplier(PROPDMGEXP), CROPDMGNUMERIC = CROPDMG * exponentToMultiplier(CROPDMGEXP))
```

## Results

### By impact on public health

Regarding the impact of weather phenomena on population's health, the NOAA data used in this research contains
two components:
- fatalities
- injuries.

Let's detect the conditions that are the most harmful in these aspects.

In this step we compute the total number of fatalities for each event type by means of the following code.
The output shows the most severe weather factors.

```{r}
byFatalities <- storms %>% select(EVTYPE, FATALITIES) %>% group_by(EVTYPE) %>%
    summarize(TOTALFATALITIES = sum(FATALITIES)) %>% arrange(desc(TOTALFATALITIES))

byFatalities[1:15,]
```

To understand the overall image, it's more convenient to have a graphical representation of the summary we've just made.
As there are too many event types in the table, we need some preprocessing to get a clear image. 
Let's plot separately only a few most important factors and group together the rest of weather conditions labeling them as 'OTHERS'.

```{r}
numPlottedFactors <- 15
byFatalitiesMostImportant <- byFatalities[1:numPlottedFactors, ]
byFatalitiesLessImportant <- byFatalities[(numPlottedFactors + 1):nrow(byFatalities), ]
byFatalitiesLessImportant <- byFatalitiesLessImportant %>% summarize(EVTYPE = "OTHERS", TOTALFATALITIES = sum(TOTALFATALITIES))
byFatalitiesFiltered <- rbind(byFatalitiesMostImportant, byFatalitiesLessImportant)
byFatalitiesFiltered$EVTYPE <- reorder(byFatalitiesFiltered$EVTYPE,
                                   -byFatalitiesFiltered$TOTALFATALITIES)
```

The actual visualization is performed this way:

```{r}
library(ggplot2)
g <- ggplot(byFatalitiesFiltered, aes(x = EVTYPE, y = TOTALFATALITIES, fill = EVTYPE)) + 
    geom_bar(width = 1, stat = "identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    labs(title = "Total fatalities by event type", x = "Event type", y = "Total fatalities")
g
```

Now exactly the same procedure is applied to the injuries data.

```{r}
byInjuries <- storms %>% select(EVTYPE, INJURIES) %>% group_by(EVTYPE) %>%
    summarize(TOTALINJURIES = sum(INJURIES)) %>% arrange(desc(TOTALINJURIES))

byInjuries[1:15,]
```

Like in the fatalities analysis, we summarize less severe conditions and place them in a category called 'OTHERS'.

```{r}
byInjuriesMostImportant <- byInjuries[1:numPlottedFactors, ]
byInjuriesLessImportant <- byInjuries[(numPlottedFactors + 1):nrow(byFatalities), ]
byInjuriesLessImportant <- byInjuriesLessImportant %>% summarize(EVTYPE = "OTHERS", TOTALINJURIES = sum(TOTALINJURIES))
byInjuriesFiltered <- rbind(byInjuriesMostImportant, byInjuriesLessImportant)
byInjuriesFiltered$EVTYPE <- reorder(byInjuriesFiltered$EVTYPE, 
                                     -byInjuriesFiltered$TOTALINJURIES)

g <- ggplot(byInjuriesFiltered, aes(x = EVTYPE, y = TOTALINJURIES, fill = EVTYPE)) + 
    geom_bar(width = 1, stat = "identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    labs(title = "Total injuries by event type", x = "Event type", y = "Total injuries")

g
```

Although there some differences between the obtained lists, in general the conditions killing many people also are the worst in terms
of injuries. One thing easy to observe from the plot is a relatively high number of injuries caused by tornados.

Let's consider the overall impact on population health as the sum of fatalities and injuries. Here is the code performing a similar analysis
and its output.

```{r}
byHealthDamage <- storms %>% select(EVTYPE, FATALITIES, INJURIES) %>% mutate(HEALTHDAMAGE = FATALITIES + INJURIES) %>%
    group_by(EVTYPE) %>% summarize(TOTALHEALTHDAMAGE = sum(HEALTHDAMAGE)) %>% arrange(desc(TOTALHEALTHDAMAGE))

byHealthDamage[1:15,]
```


### By overall economic damage

The aim of this part of the research is detecting the weather phenomena causing the largest damage to the economy.
Let's compute total damage by phenomenon type, order the totals in descending order.
After that, we can see the types corresponding to the worst impact on the economy (property and crops).

```{r}
byDamage <- storms %>% select(EVTYPE, DAMAGENUMERIC) %>% group_by(EVTYPE) %>%
    summarize(TOTALDAMAGE = sum(DAMAGENUMERIC)) %>%
    arrange(desc(TOTALDAMAGE))

byDamage[1:15, ]
```


A plot comparing damage caused by different weather conditions is built. As in the previous steps, we separate the most important factors
and label the rest as 'OTHERS'. Let's set a threshold value.
Then we group together the event types causing damage levels lower than the threshold and label them as 'OTHERS'.

```{r}
damageThreshold <- 8e9
byDamageMostImportant <- byDamage %>% filter(TOTALDAMAGE >= damageThreshold)
byDamageLessImportant <- byDamage %>% filter(TOTALDAMAGE < damageThreshold)
byDamageLessImportant <- byDamageLessImportant %>% 
    summarize(EVTYPE = "OTHERS", TOTALDAMAGE = sum(TOTALDAMAGE))
byDamageFiltered <- rbind(byDamageMostImportant, byDamageLessImportant)
byDamageFiltered$EVTYPE <- reorder(byDamageFiltered$EVTYPE,
                                   -byDamageFiltered$TOTALDAMAGE)
```

The following code builds a bar plot comparing damage by event type.

```{r}
g <- ggplot(byDamageFiltered, aes(x = EVTYPE, y = TOTALDAMAGE / 1e9, fill = EVTYPE)) + 
    geom_bar(width = 1, stat = "identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    labs(title = "Total economic damage by event type", x = "Event type", y = "Total damage, billion dollars")
g
```

### Comparison of damage to property and crops

Let's perform a similar procedure to analyze only the damage to property.

```{r}
byDamageProp <- storms %>% select(EVTYPE, PROPDMGNUMERIC) %>%
    group_by(EVTYPE) %>% summarize(TOTALDAMAGE = sum(PROPDMGNUMERIC)) %>% arrange(desc(TOTALDAMAGE))
byDamageProp[1:15,]
```

Also we need a similar summary for the damage to crops.

```{r}
byDamageCrop <- storms %>% select(EVTYPE, CROPDMGNUMERIC) %>%
    group_by(EVTYPE) %>% summarize(TOTALDAMAGE = sum(CROPDMGNUMERIC)) %>% arrange(desc(TOTALDAMAGE))
byDamageCrop[1:15,]
```

This simple analysis shows that the worst weather phenomena damaging crops are somewhat different from those causing the worst damage to property.

## Conclusion

We computed some total numbers describing the damage caused to the US by different weather phenomena during 1950-2011.
The worst phenomena turned out to be:
- Regarding fatalities: tornado, excessive heat and flash flood.
- Regarding injuries: TSTM (thunderstorm?) wind and flood.
- Regarding the overall damage to economy: flood, hurricane/typhoon and tornado.
